---
title: "fireSense"
author: "Jean Marchal"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_document: default
  pdf_document: default
---

# Overview
A landscape fire model sensitive to environmental changes (e.g. weather and land-cover).

# Download the module
```{r download module, eval = FALSE, echo = TRUE}
library(SpaDES)

moduleName <- "fireSense"

workingDirectory <- tempdir() # Location where the module will be downloaded

downloadModule(moduleName, path = workingDirectory)
```

# Usage
## Module parameters
Name|Default|Description
----|:-------|---------------------------------------------------------------------
`mapping`|`NULL`|optional named vector or list of character strings mapping one or more inputs required by the module to objects loaded in the simList environment.
`.runInitialTime`|`start(simList)`|when to start this module? By default, the start time of the simulation.
`.runInterval`|`NA`|optional. Interval between two runs of this module, expressed in units of simulation time.
`.saveInitialTime`|`NA`|optional. When to start saving output to a file.
`.saveInterval`|`NA`|optional. Interval between save events.
`.plotInitialTime`|`NA`|optional. When to start plotting.
`.plotInterval`|`NA`|optional. Interval between plot events.
|||

## Usage example
```{r module usage example, eval = FALSE}
library(SpaDES)
library(magrittr)

# Packages required by the module
library(data.table)
library(raster)

set.seed(1)

nx <- ny <- 100L
n <- nx * ny
r <- raster(nrows = ny, ncols = nx, xmn = -nx/2, xmx = nx/2, ymn = -ny/2, ymx = ny/2)

# Create a map ignition probabilities
ignitionProb <- gaussMap(r, scale = 10, var = .0001, speedup = nx/5e2, inMemory = TRUE)

# Create a map of escape probabilities
escapeProbabilities <- gaussMap(r, scale = 50, var = .01, speedup = nx/5e2, inMemory = TRUE)

#outputDir <- file.path(tempdir(), "outputs")
times <- list(start = 1, end = 100, timeunit = "year")
parameters <- list(
  fireSense_NWT = list(mapping = list(escapeProb = "escapeProbabilities") # One can use mapping to map data objects in the
                                                                      # simList environment to data required by the module.
                                                                      # Here random_ageMap is mapped to ageMap.  
  )
)

modules <- list("fireSense_NWT")
objects <- c("ignitionProb",
             "escapeProbabilities")
paths <- list(
  # cachePath = file.path(outputDir, "cache"),
  modulePath = ".."
  # inputPath = inputDir,
  # outputPath = outputDir
)

mySim <- simInit(times = times, params = parameters, modules = modules, objects = objects, paths = paths)

mySimOut <- spades(mySim)
```

# Events
- ignite fires
- determine which fires escape

# Data dependencies
## Input data
- **ignitProb**: a RasterLayer or RasterStack (time series) describing spatial variations in ignition probabilities.
- **escapeProb**: a RasterLayer or RasterStack (time series) describing spatial variations in escape probabilities.

## Output data
A umeric vector describing pixels where to start fires.
