---
title: "waterlandClassification"
author: "tmichele"
date: "28 January 2019"
output: pdf_document
---

# Overview

This is a module to define water, wetland and uplands based on Hybrid Wetland Layer from Ducks Unlimited Canada v. 2.1.
It creates a classified with 250m resolution, and projection from the Hybrid Wetland Layer using the dominant (> 50%) classification 
from DUCK's layer.

# Usage

```{r module_usage}
library(SpaDES)

setPaths(modulePath = file.path(getwd(), "modules"))
getPaths() # shows where the 4 relevant paths are

times <- list(start = 0, end = 1)

parameters <- list()
modules <- list("waterlandClassification")
objects <- list()
inputs <- list()
outputs <- list()

mySim <- simInit(times = times, params = parameters, modules = modules,
                 objects = objects)

mySimOut <- spades(mySim, debug = 2)

raster::plot(mySimOut$wetLCC)
```

## Output data

The following code can access the layer:
```{r output}

mySimOut$wetLCC

```

## Output data

Alternatively, the layer can be easily created outside of a `SpaDES` module accessible from:

```{r output}

wetLCC <- reproducible::prepInputs(url = "https://drive.google.com/open?id=1YVTcIexNk-obATw2ahrgxA6uvIlr-6xm",
                                   targetFile = "wetlandsNWT250m.tif", 
                                   destinationPath = getPaths()$output)

```

