---
title: "birdsPosthoc"
author: "tmichele"
date: "29 October 2019"
output: pdf_document
---

# Overview

Provide an overview of what the module does / how to use the module.

Module documentation should be written so that others can use your module.
This is a template for module documentation, and should be changed to reflect your module.

```{r}
typeOfSimulation <- "dynamic" #"vegetationStatic", "dynamic"
modelVersion <- "V4"# "V6" #For birds only
```

# Usage

## Saving/Uploading

```{r}
googleFoldersList <- list(
# vegetation static # Only important for birds, and we are not doing this anymore...  
vegetationStatic = list(
#   LandR.CS_fS = "1VD2DLb9Pxpah06jtmIJOMBCLot8QjgvF",
#   LandR.CS_SCFM = "1rKIBroCNPl1JmFp9qcPNf5tlxrvx0IJu",
#   LandR_fS = "1AqmRjPoueWm_H7j2iOlSgPm_lgOraL7c",
#   LandR_SCFM = "1_O-ZIJEyaOCSv2VfLMFPPHSgY4Rqp8y3"
  ),
# climate static
climateStatic = list(
  # LandR.CS_fS = "171UdUyx_-0zZXh_mHNl-uIFGHs1mzdqT",
  # LandR.CS_SCFM = "1BZkOops0CQ3rnBBY_2g38KshI5tcX5XS",
  # LandR_fS = "1SjNwvkCOd42LFTmVGLvz_G3xQyUc9NxM",
  # LandR_SCFM = "19XIS2R2XBEqCoYAEpKdILr4U6uCpdbso"
  ),
# dynamic simulation
dynamic = list(
  LandR.CS_fS = "1My_5RIOMBVyuQ7RhyiZP2Q1sHp4dlLN3",
  # LandR.CS_SCFM = "1Nuo091FmyFnrfLKVvQXjGoTN8fRUfXIJ",
  # LandR_fS = "1PmQR0SkNOrfi5fq_a16PTmsYoLw8zSJJ",
  LandR_SCFM = "1DlooFy1hYu9yOHiOOiS4uA2RrbnfxRZV")
)
```


```{r module_usage}
  library("SpaDES")
  library("future")
  library("future.apply")
  plan("sequential")
  googledrive::drive_auth("tati.micheletti@gmail.com")
  library("usefun")
# devtools::load_all("/mnt/data/Micheletti/usefun/")
  
  if (pemisc::user() %in% c("Tati", "tmichele"))
    setwd("/mnt/data/Micheletti/NWT")
  
  times <- list(start = 1, end = 1)
  Run <- "run1"
  
    inputs <- list()
  outputs <- data.frame(
    objectName = c("deltaRasters",
                   "significantChanges",
                   "pixelsSummaries",
                   "RSFlikePlot",
                   "averageInTime",
                   "averageComparison"),
    saveTime = times$end)

  doCaribou <- FALSE
  doBirds <- FALSE
  
  #  ~~~~~~~~~~~~~~~~~~~~~~~~# FOR BIRDS #~~~~~~~~~~~~~~~~~~~~~~~~
  
  if (doBirds){
      setPaths(modulePath = file.path(getwd(), "modules"),
           inputPath = file.path(getwd(), "outputs/05DEC19/"),
           outputPath = checkPath(file.path(getwd(), "outputs/05DEC19/birdResults", typeOfSimulation, Run), create = TRUE)
  )
  getPaths() # shows where the 4 relevant paths are

     species <- c("RUBL", "CAWA", "OSFL") #c("RUBL", "CAWA", "OSFL", "BBWA", "WEWP")
  parameters <- list(
    rastersPosthoc = list(patternsToRetrieveRasters = c("predicted", "Year", "tif"),
                          patternsUsedForGrouping = species,
                          years = c(seq(2011, 2100, by = 30), 2100),
                          species = species, 
                          n = 50,
                          sampleSize = 45,
                          typeOfAnalysis = paste0("birds_", typeOfSimulation),
                          uploadPlots = TRUE,
                          calculateSignificantChanges = TRUE,
                          calculateSummary = TRUE)
    )
  modules <- list("rastersPosthoc")
  objects <- list(dataFolder = list(
                  LandR.CS_fS_V6 = file.path(getPaths()$inputPath, "LandR.CS_fS",
                                             Run, paste0("birdPredictions", "V6", typeOfSimulation)), # REPLACE V6 with modelVersion
                  # LandR_fS_V4 = file.path(getPaths()$inputPath, "LandR_fS", 
                                          # Run, paste0("birdPredictions", modelVersion, typeOfSimulation)),
                  # LandR.CS_SCFM_V4 = file.path(getPaths()$inputPath, "LandR.CS_SCFM", 
                                               # Run, paste0("birdPredictions", modelVersion, typeOfSimulation)),
                  LandR_SCFM_V4 = file.path(getPaths()$inputPath, "LandR_SCFM", 
                                            Run, paste0("birdPredictions", modelVersion, typeOfSimulation))),
                  googleFolders = googleFoldersList[[typeOfSimulation]])
  
birdResults <- simInitAndSpades(times = times, params = parameters, modules = modules,
                  objects = objects, debug = 1)
  }
  
  #  ~~~~~~~~~~~~~~~~~~~~~~~~# FOR CARIBOUS #~~~~~~~~~~~~~~~~~~~~~~~~
if (doCaribou){
    setPaths(modulePath = file.path(getwd(), "modules"),
           inputPath = file.path(getwd(), "outputs/05DEC19/"),
           outputPath = checkPath(file.path(getwd(), "outputs/05DEC19/caribouResults"), create = TRUE)
  )
  getPaths() # shows where the 4 relevant paths are

  typeOfRas <- c("SelectionTaiga", "SelectionUncertainTaiga")
  parameters <- list(
    rastersPosthoc = list(patternsToRetrieveRasters = c("relativeSelection", "Year", "tif"),
                          patternsUsedForGrouping = typeOfRas,
                          years = c(2011, 2041, 2071, 2100),
                          species = typeOfRas, 
                          n = 50,
                          sampleSize = 45,
                          typeOfAnalysis = "Caribou",
                          uploadPlots = TRUE,
                          calculateSignificantChanges = FALSE,
                          calculateSummary = TRUE,
                          makeRSFLikePlot = TRUE)
    )
  modules <- list("rastersPosthoc")
  objectsCaribou <- list(dataFolder = list(
                  LandR.CS_fS = file.path(getPaths()$inputPath, "LandR.CS_fS", Run, "caribouPredictions"),
                  # LandR_fS = file.path(getPaths()$inputPath, "LandR_fS", Run, "caribouPredictions"),
                  # LandR.CS_SCFM = file.path(getPaths()$inputPath, "LandR.CS_SCFM", Run, "caribouPredictions"),
                  LandR_SCFM = file.path(getPaths()$inputPath, "LandR_SCFM", Run, "caribouPredictions")),
                  googleFolders = googleFoldersList[[typeOfSimulation]]
                  )

boo <- simInitAndSpades(times = times, params = parameters, modules = modules,
                   objects = objectsCaribou, outputs = outputs)
}

```

# Events

Describe what happens for each event type.

## Plotting

Write what is plotted.

# Data dependencies

## Input data

How to obtain input data, and a description of the data required by the module.
If `sourceURL` is specified, `downloadData("birdsPosthoc", "path/to/modules/dir")` may be sufficient.

## Output data

Description of the module outputs.

# Links to other modules

Describe any anticipated linkages to other modules.

